### 2018-02-11 submission/manuscript.Rmd
update methods section to make clearer, hopefully

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for $\alpha$-diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used. Relative risk was analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study. Relative risk significance testing utilized the chi-squared test. $\beta$-diversity differences utilized a Bray-Curtis distance matrix and PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity was first assessed for differences between controls, adenoma stage, and carcinoma stage using both linear mixed-effect models and OR. For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma. Next, all common genera were assessed for differences in OR for having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased OR based on P-value). Comparison between the full and selected models took two different approaches. In the first approach, models were trained on one study then tested on the remaining studies. This process was repeated for every study in the meta-analysis. In the second approach models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma individuals were made, the carcinoma individuals were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study. For all analysis completed feces and tissue samples were kept separate. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the OR, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings. The select models for the first approach utilized the top 5 most significantly increased and decreased ORs. For this approach, the full community models for each study were built by utilizing all genera and lowest taxonomic groups identified within that particular study. The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach. The main difference was that any OTU that taxonomically classified to one of the genera in the top 5 increased or decreased OR were included in the select model. OTU Random Forest models using the full community included all OTUs.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples with control and tumor samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing. All samples, from both tissue sample types, were analyzed together for the linear mixed-effect models with samples from the same individual corrected for. For all other analysis, not mentioned herein, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

***Assessing Important Random Forest Model Variables:***
Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained for the full models of the two different approaches used. For the first approach utilizing genus-based models, the number of times that a genus showed up in the top 10 of the training set across each study was counted. For the second approach, that utilized the OTU-based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study. Common taxa, for the OTU based models, were identified by using the lowest classification within the RDP database for each of the specific OTUs obtained from the previous counts and the number of times this classification occurred in the top 10, in each study, was recorded. Finally, for the two studies that had adenoma tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs.

>>>>

***Statistical Analysis:***
All statistical analysis after sequence processing utilized the R (v`r paste(version$major, ".", version$minor, sep = "")`) software package [@r_citation_2017]. For OTU richness, evenness, and Shannon diversity analysis, values were power transformed using the rcompanion (v`r paste(packageVersion("rcompanion")[1])`) package [@rcompanion_citation_2017] and then Z-score normalized using the car (v`r paste(packageVersion("car")[1])`) package [@car_citation_2011]. Testing for OTU richness, evenness, and Shannon diversity differences utilized linear mixed-effect models created using the lme4 (v`r paste(packageVersion("lme4"))[1]`) package [@lme4_citation_2015] to correct for study, repeat sampling of individuals (tissue only), and 16S hyper-variable region used. Odds ratios (OR) were analyzed using both the epiR (v`r paste(packageVersion("epiR"))[1]`) and metafor (v`r paste(packageVersion("metafor"))[1]`) packages [@epir_citation_2017; @metafor_citation_2010] by assessing how many individuals with and without disease were above and below the overall median value within each specific study. OR significance testing utilized the chi-squared test. Diversity differences measured by the Bray-Curtis index utilized the creation of distance matrix and testing with PERMANOVA executed with the vegan (v`r paste(packageVersion("vegan"))[1]`) package [@vegan_citation_2017]. Random Forest models were built using both the caret (v`r paste(packageVersion("caret"))[1]`) and randomForest (v`r paste(packageVersion("randomForest"))[1]`) packages [@caret_citation_2017; @randomforest_citation_2002]. Power analysis and estimations were made using the pwr (v`r paste(packageVersion("pwr"))[1]`) and statmod (v`r paste(packageVersion("statmod"))[1]`) packages [@pwr_citation_2017; @statmod_citation_2016]. All figures were created using both ggplot2 (v`r paste(packageVersion("ggplot2"))[1]`) and gridExtra (v`r paste(packageVersion("gridExtra"))[1]`) packages [@ggplot2_citation_2009; @gridextra_citation_2017].

***Study Analysis Overview:***
OTU richness, evenness, and Shannon diversity were first assessed for differences between controls, adenoma tumors, and carcinoma turmors using both linear mixed-effect models and ORs. For each individual study the Bray-Curtis index was used to assess differences between control-adenoma and control-carcinoma individuals. Next, all common genera were assessed for differences in ORs for individuals having an adenoma or carcinoma and ranked based on P-value. We then built Random Forest models based on the full or selected community (the top 5 increased and top 5 decreased ORs based on P-value). Comparison between the full and selected models took two different approaches. In the first approach, models were trained on one study then tested on the remaining studies using genera-based relative abundances. This process was repeated for every study in the meta-analysis. In the second approach, models were built using OTU level data and a 10-fold CV over 100 different iterations, based on random 80/20 splitting of the data, was used to generate a range of expected AUCs. For these OTU-based models, the selected model included all OTUs that had a taxonomic classification to a taxa in the top 5 increased and top 5 decreased OR analysis based on P-value. Finally, the power of each study was assessed for an effect size ranging from 1% to 30% and an estimated sample size, for these effect sizes, was generated based on 80% power. For comparisons in which only control versus adenoma individuals were made, the carcinoma individuals were excluded from each respective study. Similarly, for comparisons in which control versus carcinoma individuals were made the adenoma individuals were excluded from each respective study. For all analysis completed fecal and tissue samples were kept separate. Within the tissue groups the data were further divided between samples from the same individual (matched) and those from different individuals (unmatched).

***Obtaining Genera Relative Abundance and Selected Models:***
For the genera analysis of the ORs, OTUs were added together based on the genus or lowest available taxonomic classification level and the total average counts, for 100 different subsamplings was obtained. The select models for the first Random Forest model approach utilized the top 5 most significantly increased and decreased ORs. For this approach, the full community models for each study were built by utilizing all genera and lowest taxonomic groups identified within that particular study. The select models for the second approach, that used OTU-based Random Forest models, utilized a similar method as the first approach. The main difference was that any OTU that taxonomically classified to one of the taxa in the top 5 increased or decreased OR analysis were included in the select model. OTU Random Forest models, using the full community, included all OTUs that did not have near zero variance within that specific study.

***Matched versus Unmatched Tissue Samples:***
In general, tissue samples with control and tumor samples from different individuals were classified as unmatched while samples that belonged to the same individual were classified as matched. Studies with matched data included Burns, Dejea, Geng, and Lu while those with unmatched data were from Burns, Flemer, Chen, and Sanapareddy. For some studies samples became unmatched when a corresponding matched sample did not make it through sequence processing. All samples, from both matched and unmatched tissue samples, were analyzed together for the linear mixed-effect models with samples from the same individual being corrected for. All other analysis, where it is not specified explicitly, matched and unmatched samples were analyzed separately using the statistical approaches mentioned in the Statistical Analysis section.

***Assessing Important Random Forest Model Variables:***
Using Mean Decrease in Accuracy (MDA) the top 10 most important variables to the Random Forest model were obtained for the full models of the two different approaches used. For the first approach utilizing genus-based models, the number of times that a specific taxa showed up in the top 10 of the training set across each study was counted. For the second approach, that utilized the OTU-based models, the medians for each OTU across 100 different 80/20 splits of the data was generated and the top 10 OTUs then counted for each study. Common taxa were then identified by using the lowest classification for each of the specific OTUs obtained from these counts and the number of times this classification occurred across the top 10 of each study was recorded. Finally, the two studies that had adenoma tumor tissue (Lu and Flemer) were equally divided between matched and unmatched studies and were grouped together for the counting of the top 10 genera and OTUs for both Random Forest approaches.

### 2018-02-11 submission/manuscript.tex
update methods section to make clearer, hopefully

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}39{]} and
then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
for \(\alpha\)-diversity differences utilized linear mixed-effect models
created using the lme4 (v1.1.15) package {[}41{]} to correct for study,
repeat sampling of individuals (tissue only), and 16S hyper-variable
region used. Relative risk was analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages {[}42,43{]} by assessing how many
individuals with and without disease were above and below the overall
median value within each specific study. Relative risk significance
testing utilized the chi-squared test. \(\beta\)-diversity differences
utilized a Bray-Curtis distance matrix and PERMANOVA executed with the
vegan (v2.4.5) package {[}44{]}. Random Forest models were built using
both the caret (v6.0.78) and randomForest (v4.6.12) packages
{[}45,46{]}. Power analysis and estimations were made using the pwr
(v1.2.1) and statmod (v1.4.30) packages {[}47,48{]}. All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages
{[}49,50{]}.

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity was first assessed for differences between controls,
adenoma stage, and carcinoma stage using both linear mixed-effect models
and OR. For each individual study the Bray-Curtis index was used to
assess differences between control-adenoma and control-carcinoma. Next,
all common genera were assessed for differences in OR for having an
adenoma or carcinoma and ranked based on P-value. We then built Random
Forest models based on the full or selected community (the top 5
increased and top 5 decreased OR based on P-value). Comparison between
the full and selected models took two different approaches. In the first
approach, models were trained on one study then tested on the remaining
studies. This process was repeated for every study in the meta-analysis.
In the second approach models were built using OTU level data and a
10-fold CV over 100 different iterations, based on random 80/20
splitting of the data, was used to generate a range of expected AUCs.
For these OTU-based models, the selected model included all OTUs that
had a taxonomic classification to a taxa in the top 5 increased and top
5 decreased OR based on P-value. Finally, the power of each study was
assessed for an effect size ranging from 1\% to 30\% and an estimated
sample size, for these effect sizes, was generated based on 80\% power.
For comparisons in which only control versus adenoma individuals were
made, the carcinoma individuals were excluded from each respective
study. Similarly, for comparisons in which control versus carcinoma
individuals were made the adenoma individuals were excluded from each
respective study. For all analysis completed feces and tissue samples
were kept separate. Within the tissue groups the data were further
divided between samples from the same individual (matched) and those
from different individuals (unmatched).

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the OR, OTUs were added together based on the
genus or lowest available taxonomic classification level and the total
average counts, for 100 different subsamplings. The select models for
the first approach utilized the top 5 most significantly increased and
decreased ORs. For this approach, the full community models for each
study were built by utilizing all genera and lowest taxonomic groups
identified within that particular study. The select models for the
second approach, that used OTU-based Random Forest models, utilized a
similar method as the first approach. The main difference was that any
OTU that taxonomically classified to one of the genera in the top 5
increased or decreased OR were included in the select model. OTU Random
Forest models using the full community included all OTUs.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples with control and tumor samples from different individuals
were classified as unmatched while samples that belonged to the same
individual were classified as matched. Studies with matched data
included Burns, Dejea, Geng, and Lu while those with unmatched data were
from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
became unmatched when a corresponding matched sample did not make it
through sequence processing. All samples, from both tissue sample types,
were analyzed together for the linear mixed-effect models with samples
from the same individual corrected for. For all other analysis, not
mentioned herein, matched and unmatched samples were analyzed separately
using the statistical approaches mentioned in the Statistical Analysis
section.

\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
Mean Decrease in Accuracy (MDA) the top 10 most important variables to
the Random Forest model were obtained for the full models of the two
different approaches used. For the first approach utilizing genus-based
models, the number of times that a genus showed up in the top 10 of the
training set across each study was counted. For the second approach,
that utilized the OTU-based models, the medians for each OTU across 100
different 80/20 splits of the data was generated and the top 10 OTUs
then counted for each study. Common taxa, for the OTU based models, were
identified by using the lowest classification within the RDP database
for each of the specific OTUs obtained from the previous counts and the
number of times this classification occurred in the top 10, in each
study, was recorded. Finally, for the two studies that had adenoma
tissue (Lu and Flemer) were equally divided between matched and
unmatched studies and were grouped together for the counting of the top
10 genera and OTUs.

>>>>

\textbf{\emph{Statistical Analysis:}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package {[}38{]}.
For OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package {[}39{]} and
then Z-score normalized using the car (v2.1.6) package {[}40{]}. Testing
for OTU richness, evenness, and Shannon diversity differences utilized
linear mixed-effect models created using the lme4 (v1.1.15) package
{[}41{]} to correct for study, repeat sampling of individuals (tissue
only), and 16S hyper-variable region used. Odds ratios (OR) were
analyzed using both the epiR (v0.9.93) and metafor (v2.0.0) packages
{[}42,43{]} by assessing how many individuals with and without disease
were above and below the overall median value within each specific
study. OR significance testing utilized the chi-squared test. Diversity
differences measured by the Bray-Curtis index utilized the creation of
distance matrix and testing with PERMANOVA executed with the vegan
(v2.4.5) package {[}44{]}. Random Forest models were built using both
the caret (v6.0.78) and randomForest (v4.6.12) packages {[}45,46{]}.
Power analysis and estimations were made using the pwr (v1.2.1) and
statmod (v1.4.30) packages {[}47,48{]}. All figures were created using
both ggplot2 (v2.2.1) and gridExtra (v2.3) packages {[}49,50{]}.

\textbf{\emph{Study Analysis Overview:}} OTU richness, evenness, and
Shannon diversity were first assessed for differences between controls,
adenoma tumors, and carcinoma turmors using both linear mixed-effect
models and ORs. For each individual study the Bray-Curtis index was used
to assess differences between control-adenoma and control-carcinoma
individuals. Next, all common genera were assessed for differences in
ORs for individuals having an adenoma or carcinoma and ranked based on
P-value. We then built Random Forest models based on the full or
selected community (the top 5 increased and top 5 decreased ORs based on
P-value). Comparison between the full and selected models took two
different approaches. In the first approach, models were trained on one
study then tested on the remaining studies using genera-based relative
abundances. This process was repeated for every study in the
meta-analysis. In the second approach, models were built using OTU level
data and a 10-fold CV over 100 different iterations, based on random
80/20 splitting of the data, was used to generate a range of expected
AUCs. For these OTU-based models, the selected model included all OTUs
that had a taxonomic classification to a taxa in the top 5 increased and
top 5 decreased OR analysis based on P-value. Finally, the power of each
study was assessed for an effect size ranging from 1\% to 30\% and an
estimated sample size, for these effect sizes, was generated based on
80\% power. For comparisons in which only control versus adenoma
individuals were made, the carcinoma individuals were excluded from each
respective study. Similarly, for comparisons in which control versus
carcinoma individuals were made the adenoma individuals were excluded
from each respective study. For all analysis completed fecal and tissue
samples were kept separate. Within the tissue groups the data were
further divided between samples from the same individual (matched) and
those from different individuals (unmatched).

\textbf{\emph{Obtaining Genera Relative Abundance and Selected Models:}}
For the genera analysis of the ORs, OTUs were added together based on
the genus or lowest available taxonomic classification level and the
total average counts, for 100 different subsamplings was obtained. The
select models for the first Random Forest model approach utilized the
top 5 most significantly increased and decreased ORs. For this approach,
the full community models for each study were built by utilizing all
genera and lowest taxonomic groups identified within that particular
study. The select models for the second approach, that used OTU-based
Random Forest models, utilized a similar method as the first approach.
The main difference was that any OTU that taxonomically classified to
one of the taxa in the top 5 increased or decreased OR analysis were
included in the select model. OTU Random Forest models, using the full
community, included all OTUs that did not have near zero variance within
that specific study.

\textbf{\emph{Matched versus Unmatched Tissue Samples:}} In general,
tissue samples with control and tumor samples from different individuals
were classified as unmatched while samples that belonged to the same
individual were classified as matched. Studies with matched data
included Burns, Dejea, Geng, and Lu while those with unmatched data were
from Burns, Flemer, Chen, and Sanapareddy. For some studies samples
became unmatched when a corresponding matched sample did not make it
through sequence processing. All samples, from both matched and
unmatched tissue samples, were analyzed together for the linear
mixed-effect models with samples from the same individual being
corrected for. All other analysis, where it is not specified explicitly,
matched and unmatched samples were analyzed separately using the
statistical approaches mentioned in the Statistical Analysis section.

\textbf{\emph{Assessing Important Random Forest Model Variables:}} Using
Mean Decrease in Accuracy (MDA) the top 10 most important variables to
the Random Forest model were obtained for the full models of the two
different approaches used. For the first approach utilizing genus-based
models, the number of times that a specific taxa showed up in the top 10
of the training set across each study was counted. For the second
approach, that utilized the OTU-based models, the medians for each OTU
across 100 different 80/20 splits of the data was generated and the top
10 OTUs then counted for each study. Common taxa were then identified by
using the lowest classification for each of the specific OTUs obtained
from these counts and the number of times this classification occurred
across the top 10 of each study was recorded. Finally, the two studies
that had adenoma tumor tissue (Lu and Flemer) were equally divided
between matched and unmatched studies and were grouped together for the
counting of the top 10 genera and OTUs for both Random Forest
approaches.
